---
title: "ggplot Package"
author: "23BMAT019"
date: "2023-10-27"
output: 
  html_document: 
    toc: yes
    highlight: kate
    theme: spacelab
---

```{r}
library("dplyr")
library("ggplot2")
library("ggplot2movies")
library("Hmisc")
library("tidyverse")
```

Key Points:

- data
- mapping (aesthetics [x v/s y])
- geometric interpretation
- statistics
- facet
- co-ordinate space
- label
- theme

`+` is used to join the commands

# Scatter Plot and Regression

## Example (BOD: Biochemical Oxygen Demand)
```{r}
View(BOD)

#ggplot(BOD, aes(Time, demand))
BOD %>%
  ggplot(aes(Time, demand))+
  geom_line(size=2, color="maroon")+
  geom_point(size=4, color="darkblue")+
  theme_classic()
```

## Example (CO2)
```{r}
View(CO2)

CO2 %>%
  ggplot(aes(conc, uptake, color=Treatment))+
  #ggplot(aes(conc, uptake, color=Plant))+
  geom_point()+
  geom_smooth(method=lm, se=F)+
  facet_wrap(~Type)+ # Preferably use factor columns
  labs(title="Conc of CO2")+
  theme_dark()

# lm: linear model
# se: standard error
```

## Example (mpg)
```{r}
View(mpg)
#1 Plot something for the data where `cty` < 20
mpg %>% filter(cty < 20) %>%
  ggplot(aes(displ, cty))+
  geom_point(aes(color=drv))+
  geom_smooth(method=lm, se=F, color="maroon")+
  facet_wrap(~cyl)+
  labs(x="Engine Size", y="City miles per Gallon", title="Fuel Efficiency")+
  theme_bw()
```

# Boxplot

## Example (CO2)
```{r}
CO2 %>% ggplot(aes(Treatment, uptake))+
  geom_boxplot()+
  geom_point(aes(color=Plant, size=conc))+
  facet_wrap(~Plant)+
  labs(title="Boxplot")+
  coord_flip()+
  theme_bw()
```

# Example (chickwts)
```{r}
chickwts %>% ggplot(aes(x=reorder(feed, weight), y=weight, fill=feed))+
  geom_boxplot(outlier.colour = "red", alpha=1.9, outlier.shape = 3, outlier.size = 3)+# to hide the outliers, use outliers.shape=NA
  #coord_flip()+ # for horizontal boxplots, use coordinate flip
  geom_jitter(colour="green", size=0.3)+ # to get the points, use geom_point() but those points overlap each other. geom_jitter() is used to avert overlapping
  stat_boxplot(geom="errorbar", width=0.2)+ # to add the lines at end of each boxplot line
  stat_summary(fun=mean, geom="point", col="maroon", size=2)+ # to add a mean in form of point in maroon of size 2
  #stat_summary(fun.data=mean_cl_normal, geom="linerange", col="yellow")+
  labs(title="Box Plot")+
  labs(subtitle="with additional conditions")+
  theme_dark()

# we can fill the colour inside geom_boxplot(): geom_boxplot(fill="green")

# higher the alpha in ]0,1[, lower the transparency

# to reorder with a specific condition, use reordr in aes(x=(a,b)) read as 'on x-axis reorder a according to b' (for descending use -b)
```

# Histogram

## Example (msleep)
```{r}
msleep %>% ggplot(aes(x=awake))+ # y will be the frequency
  geom_histogram(binwidth = 1, # size of each bin
                 aes(y=..density..), # density function
                 color="green",
                 fill="purple")+
  labs(x="Total Sleep", y=NULL,
       title="Histogram",
       subtitle="Of Total Sleep")+
  theme_dark()
```

## Example (Plotting the density function)
```{r}
data = data.frame(x=rnorm(1E+5))
data %>% ggplot(aes(x))+
  geom_histogram(aes(y=..density..), color="green", fill="lightblue")+
  stat_function(fun=dnorm, 
                args=list(mean=mean(data$x),
                          sd=sd(data$x)),
                color="red",
                size=3)+
  labs(title="Histogram and Density Function")+
  theme_bw()


# dnorm(a) density function of normal distribution

```


# Bar plots

## Single Category Bar Plot
```{r}
msleep %>% drop_na(vore) %>% ggplot(aes(x=vore))+
  geom_bar(fill="darkblue", color="red")+
  coord_flip()+
  labs(title="Single Category Barplot",
       subtitle="msleep")+
  theme_bw()
```

## Two Category Bar Plot
```{r}
starwars %>% drop_na(hair_color) %>% drop_na(species) %>% drop_na(homeworld) %>% filter(hair_color == "black" | hair_color == "brown") %>% ggplot(aes(hair_color, color=homeworld, fill=gender))+
  facet_wrap(~species)+
  geom_bar(position="dodge")+ # for grouped bar plot use position="dodge"
  labs(x="Hair Colour", y="Count",
       title="Hair Color for Different Species")
```
