---
title: "Quantile-Quantile Plots"
author: "23BMAT019"
date: "2023-12-01"
output: html_document
---

**QQPlots are used to check if the data is normallly distributed or not.**

```{r}
data <- c(3, 5, 7, 5, 3,
          2, 6, 8, 5, 6, 
          9, 4, 5, 7, 3, 4)
qqnorm(data) # Quantile Plot
#qqline(data) # Quantile Line
qqnorm(data,
       pch = 20, col = "maroon",
       main = "QQ Plot of Example Data",
       xlab = "Values",
       ylab = "Quantities for data")
qqline(data,
       lwd = 2,
       lty = 3,
       col = "cornflowerblue")
```

**If most of the points lie on the line (dense around the line), then the data is normally distributed.**

Lesser the fluctuations, the more it approaches to the Normal Distribution.

## Symmetric with fat tails
```{r}

```

**Skewness** as a measure of distortion.

```{r}
norm_data <- rnorm(50, mean = 1, sd = 4)
qqnorm(norm_data,
       pch = 21, cex = 1.75, col = "orange3",
       main = "QQ Plot of Normal Dist Data",
       xlab = "Values",
       ylab = "Quantities for data")
qqline(norm_data,
       lwd = 2,
       lty = 6,
       col = "darkred")
```

Observe how most of the points are clustered around x=1

# Comparison of Distributions

```{r}
attach(mtcars)
qqnorm(cyl,
       main = "QQ-Plot for `disp` in `mtcars`",
       col = "blueviolet",
       pch = 21)
qqline(cyl, col = "maroon")
detach(mtcars)
```

# Class Questions
```{r}
urn <- runif(100, 1, 2)
nrn <- rnorm(100, 5, 1)
qqp <- qqplot(urn, nrn)
qqp
abline(lm(qqp$y ~ qqp$x))
```
