---
title: "readxl and openxlsx"
author: "You"
date: "2023-11-07"
output: 
  html_document: 
    toc: yes
    highlight: kate
    theme: spacelab
---

```{r}
# Add packages to library
library(readxl)
library(openxlsx)
```

# Functions with Syntax

1. `write`.`xlsx`(`dataframe`, "`filename.xlsx`")
  
    - Writes the `dataframe` to a file named `filename.xlsx`.
    - A Workbook object is returned
    - Please note that this function is only for .xlsx files

2. `read`.`xlsx`("`filename.xlsx`")

    - Reads `filename.xlsx` 
    - A data frame is returned

3. For Vectors
    a. `write`(`data`, `file`=`filename.ext`, `ncolumns`=if(is.character(`data`)) 1 else 5, `sep`=" ")
    
    b. `cat`(`data`, `file`=`filename.txt`, `sep`=" ")

4. For 2-D Data Structures
    a. `write`.`csv`(`data`, `file`="`filename.txt`")
    b. `write`.`table`(`data`, `file`="`filename.txt`")

5. For Lists (explore)
    a. `dget`()
    b. `dput`()
    c. `lapply`(`data`, `fun`)
    d. `sapply`(`data`, `fun`)

6. `apply`(`data`, `margin`, `fun`)
    
    - `margin` = 1 (row) or 2 (column)
    - `fun`: any function

**NOTE**:

- `is`.`<class>`(`data`)

    - Checks if the `data` is of the class `<class>`.
    - E.g. `is.numeric(data)` checks if data is numeric

- `as`.`class`(`data`)

    Converts the class of `data` to `<class>`

`read`  - **Extract** data

`write` - **Modify**  data

# Converting a Data Frame to Excel Workbook

```{r}
getwd()

head(mtcars)
class(mtcars)

# To store a Workbook object in cars.
cars <- write.xlsx(mtcars, "cars.xlsx")
## The workbook object converts `mtcars` 
## into an Excel Workbook named `cars.xlsx`

cars # Workbook object

my_wbook <- write.xlsx(iris, "23bmat019.xlsx")
my_wbook
```

# (Reading Files into R) Converting an Excel Workbook to a Data Frame 

```{r}
my_df <- read.xlsx("cars.xlsx")
head(my_df)
```


# Reading and Writing `.csv` files

```{r}
# Writing `iris` to `my_iris.csv`
my_iris <- write.csv(iris, "my_iris.csv")
my_iris # NULL

# Reading `my_iris.csv`
my_df2 <- read.csv("my_iris.csv")
head(my_df2)
```

# Writing Vectors into the Disk

```{r}
x <- c("a", "b", "c") # character data
write(x, file="my_vectx.txt")
write(x, file="my_vectx.csv") # doesn't require packages

y <- letters # character data
write(y, file="my_vecty.txt") # 1 element per row

z <- 1:35 # numeric data; more than 5 elements
write(z, file="my_vectz.txt") # 5 elements per row

write(z, file="my_vectz.csv", sep=",")
# without sep="," 5 elements are clubbed within a cell separated by spaces

# *(refer to syntax; ncolumns)
# if the class of data is character, the number of columns will be 1
# else 5 for any other class

p <- c(3, 5, 8, 9, 5, 2)
cat(p, file="my_vectp_cat.txt") # all elements in a single row
cat(p, file="my_vectp_cat.csv", sep=",")
write(p, file="my_vectp_write.txt")
```


# Writing Matrices and Data Frames into the Disk

```{r}
### Matrices
a <- matrix(1:9, 3)

# column names will be generated
# write.csv() has "," as default separator
write.csv(a, "my_matrix_csv.txt")
write.csv(a, "my_matrix_csv.csv")

# write.table() has " " as default separator
write.table(a, "my_matrix_table.txt", sep=",")
write.table(a, "my_matrix_table.csv", sep=",")

### Data Frames
write.csv(iris, "my_df_csv.txt")
write.csv(iris, "my_df_csv.csv")

# write.table() has " " as default separator
write.table(iris, "my_df_table.txt", sep=",")
write.table(iris, "my_df_table.csv", sep=",")
```

# Writing Lists into the Disk (Homework)

# Summary Statistics

## For Vector Objects (previously done)

## For Matrix Objects

```{r}
# R treats matrices like vectors

a1 = round(runif(9, 2, 89))
m1 <- matrix(a1, 3, 3,
             dimnames = list(
               c("Jan", "Feb", "Mar"),
               c("Day1", "Day2", "Day3")
             ))
m1
length(m1)
dim(m1)
nrow(m1)
ncol(m1)

sum(m1) # sum of all matrix entries
rowSums(m1) # sums of each row
colSums(m1) # sums of each column

mean(m1) # mean of all matrix entries
rowMeans(m1) # means of each row
colMeans(m1) # means of each column

var(m1) # variance of all matrix entries
# there is no rowVar and colVar command,
# instead apply() command may be used
# it can also be used for sums and means
```

### `apply()` command
```{r}
m2 = matrix(round(runif(9, 1, 99)),
            3, 3,
            dimnames = list(
              month.name[1:3],
              c("Day1", "Day2", "Day3")
            ))
m2

# the following pairs of lines are equivalent
apply(m2, 1, sum)
rowSums(m2)

apply(m2, 2, mean)
colMeans(m2)

# this will help us to map other functions on matrix dimensions
apply(m2, 1, sd) # sd of each row
apply(m2, 2, var) # variance of each column
apply(m2, 2, length) # length of each column
```

## For Data Frames (explore)

## For Lists

`lapply`() and `sapply`() will be used

```{r}
l1 = list(
  1:3,
  letters[1:5],
  c(3, 0, 6, -2, 2)
)
l1

lapply(l1, max) # output in list
sapply(l1, max) # output in vector
```


# Class Exercises

Q.1 Create a \(3 \times 3\) matrix and - 
```{r}
# initialise matrix for questions
my_matrix <- matrix(round(runif(9, 1, 99)),
                    3, 3,
                    dimnames = list(
                      LETTERS[1:3],
                      c("Week1", "Week2", "Week3")
                    ))
my_matrix
```

a. Find the variance of first and third column

```{r}
# Approaches (2):
# 1. Extract first and third columns and calculate variance.
var(my_matrix[,1])
var(my_matrix[,3])

# 2. Calculate variance for all columns and extract the first and third.
colVars = apply(my_matrix, 2, var)
colVars[1]
colVars[3]
```


b. Find the maximum of second and third row
```{r}
rowMax = apply(my_matrix, 1, max)
rowMax[1]
rowMax[3]
```

c. Find the median of first and second column
```{r}
colMedians = apply(my_matrix, 2, median)
colMedians[1]
colMedians[2]
```

Q.2 Get the first 10 rows of `mtcars`
```{r}
my_df <- head(mtcars, 10)
```

a. Find the median of each column
```{r}
apply(my_df, 2, median)
```

